---
title: "Post hoc PReMiuM Analysis"
author: "Dave Hiltbrand"
date: "4/24/2018"
output: 
  md_document: 
    variant: markdown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Post hoc Analysis

After running the G2F hybrid/weather data through the PReMiuM package there is a large amount of output.  There are also many hybrids so a further analysis is composed to try and answer a few high level questions:

  1.  Which hybrids have profile-clusters that are most different from each other (i.e. have the most g x e)?
  2.  Which weather variables are most important for making profile-clusters that differ across hybrids, i.e. which covariates contribute the most to gxe?
  3.  Which hybrid-profile clusters define the most similar fields (i.e. which field sites are the most redundant)?
  4.  Which hybrids diagnose specific envirotype groups?


```{r load_libraries}
library(amap)
library(cluster)
library(dendextend)
library(colorspace)
library(tidyverse)
library(lubridate)
library(tictoc)
library(dynamicTreeCut)
```

#### Loading data

Here load the original dataset run through Premium.  Get the output with the clustering numbers from Premium.  Make a new dataframe with the cluster associations.

```{r load_data}
setwd("~/GitHub/EnviroTyping/")

hybridByWeek <- read_rds("data/interim/G2F_Hybrid/hybrid_by_week_cleaned_weather.rds")


val <- grep("Min|Max",names(hybridByWeek))
numericVars <- names(hybridByWeek[val])[vapply(hybridByWeek[val], function(x) var(x) != 0, logical(1))]
hybridByWeekSubset <- hybridByWeek %>% select(Hyb=Pedi, Yield, Week, numericVars)
riskProfObj <- read_rds("/data/interim/G2F_Hybrid/hyb_wks3-5_by_week_1k/output/riskProfObj.rda")
weeks3_5 <- hybridByWeek %>% filter(Week >= isoweek(Planted) + 3 & Week <= isoweek(Planted) +5) %>% select(Exp,Pedi,Yield, numericVars)
weeks3_5Clust <- data.frame(optClus = as.factor(riskProfObj$riskProfClusObj$clustering), weeks3_5)
```


#### Agglomerative Hierarchical Clustering (AGNES)

As stated before there are too many hybrid labels within each cluster to make strong qualitative conclusions.  An additional step of clustering is performed using aggolmerative hierarchical clustering to further break down the data.  The _amap_ package function _hcluster_ is used instead of the standard _hclust_ as it allows for parallel processing. Several methods were also looked into when it came time to prune the dendrogram produced by AGNES.  A group at UCLA created an R package call _dynamicTreeCut_ which uses a novel way of pruning trees by accounting for nested clusters via several parameters that give the more control than the typical fixed height prune.  With a new group variable from the dynamic tree cutting and the cluster variable from the premium output, filtering the hybrid weather profiles becomes possible. 


```{r agnes}
cores <- parallel::detectCores() - 1  # Save a core for RStudio
tic()
posthocGroup <- hcluster(weeks3_5Clust[,-2], method = "euclidean", link = "ward", nbproc = cores)
toc()  # 26.264 sec elapsed

tic()
dynamicCutree <- cutreeDynamic(posthocGroup, distM = as.matrix(dist), deepSplit = 4, verbose = 4) 
toc  # 55.688 sec elapsed
n_distinct(dynamicCutree)  # 14
weeks3_5Groups <- data.frame(group = dynamicCutree, weeks3_5Clust)
```


#### Plotting the Results

Developing the workflow allows for visual indicators to drive the decisions.  The y-value, yield, was first plotted across the cluster for each group to look for any initial patterns.  In Figure 5 the same clusters can be seen in groups 1 & 6 or 2 & 5 but the yields are nearly halved in each comparison.  Second the weather profiles for each group were plotted to look for areas of separation.  Again, the goal is to find a hybrid that is experiencing a crossover interaction so hybrids that occur in different locations (i.e. different weather profiles) with varying yields.  There are some clear differences showcased between the group comparisons above in Figure 6.

```{r plotting_results}
l = list()
for (i in 1:11){
    #print({
    tmp <- weeks3_5DynamicTree %>% filter(dynamicCutree == i)
    p <-  ggplot(tmp,aes(optClus, Yield)) + geom_bar(aes(fill = optClus), stat = "summary", fun.y = "mean", width = .5) +
    scale_fill_viridis(discrete = TRUE) + labs(title = paste("Cluster by Yield for Group",i,sep = " "), x = "Envirotyping Cluster", y = "Yield [bu/acre]" ) +
    theme_bw()+ guides(fill=FALSE)
    l[[i]] <- p
    #})
}

grid.arrange(l[[1]],l[[2]],l[[3]],l[[4]],l[[5]],l[[6]],l[[7]],l[[8]],l[[9]],l[[10]],l[[11]])

png("yieldByClusterPlot.png", width = 1280, height = 1080)
ggplot(weeks3_5DynamicTree, aes(x = optClus, y = Yield)) + geom_bar(aes(fill = optClus), stat = "summary", fun.y = "mean")+
    facet_wrap(~ dynamicCutree, scales = "free_y", ncol = 4) +
    scale_fill_viridis(discrete = TRUE) + labs(title = "Cluster by Yield for Post hoc Grouping", x = "Envirotyping Cluster", y = "Yield [bu/acre]" ) +
    theme_bw()+ guides(fill=FALSE)
dev.off()

pdf("yieldByClusterPlot.pdf", paper = "a4r")
for (i in 1:14){
    print({
    tmp <- weeks3_5Groups %>% filter(group == i)
ggplot(gather(tmp,key,value, -c(group,optClus,Pedi)),aes(optClus,value)) + geom_boxplot(aes(fill=optClus)) +
    facet_wrap(~ key, scales="free_y", ncol = 5)+
    scale_fill_viridis(discrete = TRUE)+ ggtitle(paste("Cluster by Yield for Post hoc Group",i,sep = " ")) +
    theme_bw()+
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
    })
}
dev.off()
```


#### Summary of Important Findings

A table of minimum, median, and maximum values by group in Table 4 allows closer inspection to see which weather variables are driving the substantial differences between yields, Table 5.

```{r}
groupsWithHybRanks <- left_join(weeks3_5Groups,rank, by=c("Exp","Pedi"))

group2clus9 <- groupsWithHybRanks %>% filter(group==2, optClus==9) %>% select(Pedi, Rank, Yield)
group2clus13 <- groupsWithHybRanks %>% filter(group==2, optClus==13) %>% select(Pedi, Rank, Yield)
group5clus9 <- groupsWithHybRanks %>% filter(group==5, optClus==9) %>% select(Pedi, Rank, Yield)
group5clus13 <- groupsWithHybRanks %>% filter(group==5, optClus==13) %>% select(Pedi, Rank, Yield)
match(group2clus9$Pedi,group5clus9$Pedi)
match(group2clus13$Pedi,group5clus13$Pedi)

group2 <- groupsWithHybRanks %>% filter(group==2) %>% select(Exp,optClus, Pedi, Rank, Yield)
table(group2$optClus,group2$Exp)
group5 <- groupsWithHybRanks %>% filter(group==5) %>% select(Exp, optClus, Pedi, Rank, Yield)
table(group5$optClus,group5$Exp)

group2clus5 <- groupsWithHybRanks %>% filter(group==2, optClus==5) %>% select(Exp,Pedi, Rank, Yield)
group2clus11 <- groupsWithHybRanks %>% filter(group==2, optClus==11) %>% select(Exp,Pedi, Rank, Yield)
group5clus5 <- groupsWithHybRanks %>% filter(group==5, optClus==5) %>% select(Exp,Pedi, Rank, Yield)
group5clus11 <- groupsWithHybRanks %>% filter(group==5, optClus==11) %>% select(Exp,Pedi, Rank, Yield)
match(group5clus5$Pedi,group2clus5$Pedi)
match(group2clus11$Pedi,group5clus11$Pedi)

group1 <- groupsWithHybRanks %>% filter(group==1) %>% select(Exp, optClus, Pedi, Rank, Yield)
table(group1$optClus, group1$Exp)
group6 <- groupsWithHybRanks %>% filter(group==6) %>% select(Exp, optClus, Pedi, Rank, Yield)
table(group6$optClus,group6$Exp)
group1clus1 <- groupsWithHybRanks %>% filter(group==1, optClus==1) %>% select(Exp, Pedi, Rank, Yield)
group6clus1 <- groupsWithHybRanks %>% filter(group==6, optClus==1) %>% select(Exp,Pedi, Rank, Yield)
match(group1clus1$Pedi,group6clus1$Pedi)
```

